---
title: "Using gtscales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gtscales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

```{r setup}
library(gt)
library(gtscales)
library(scales)

big_number_labels <- label_number(scale_cut = cut_short_scale())
date_labels <- label_date('%b %d')
```

`gtscales` adds matched legends to color-encoded `gt` tables. The package is built around two common needs:

1. Color a table column and attach a matching legend in one step.
2. Reuse a scale definition across multiple tables, placements, or output formats.

The package supports continuous, diverging, binned, quantile, and discrete scales. It is designed to work naturally with `scales` helpers such as label functions, break functions, transform specifications, and palette functions.

## One-step helpers

For most work, the `gtscale_data_color_*()` helpers are the fastest path. They call `gt::data_color()` and then attach a matching legend.

```{r}
exibble |>
  gt() |>
  gtscale_data_color_continuous(
    column = num,
    palette = c('#A0442C', 'white', '#0063B1'),
    labels = big_number_labels,
    width = '220px',
    title = 'Numeric scale'
  )
```

Binned scales are useful when you want fixed intervals rather than a smooth gradient.

```{r}
exibble |>
  gt() |>
  gtscale_data_color_bins(
    column = currency,
    palette = c('#f7fbff', '#08306b'),
    bins = c(0, 10, 100, 1000, 10000, 70000),
    title = 'Currency bins'
  )
```

Quantile scales instead split the data into equally sized groups.

```{r}
exibble |>
  gt() |>
  gtscale_data_color_quantiles(
    column = num,
    palette = c('#fdd49e', '#fdbb84', '#ef6548', '#990000'),
    quantiles = 4,
    labels = big_number_labels,
    width = '220px',
    title = 'Quartiles'
  )
```

Discrete legends are useful when colors encode categories rather than ordered values.

```{r}
data.frame(
  district = c('A', 'B', 'C', 'D'),
  status = c('Safe D', 'Toss-up', 'Lean R', 'Safe R'),
  margin = c(18, 2, -6, -21)
) |>
  gt() |>
  gtscale_data_color_discrete(
    column = status,
    values = c('#2166ac', '#f7f7f7', '#ef8a62', '#b2182b'),
    labels = c('Safe D', 'Toss-up', 'Lean R', 'Safe R'),
    title = 'Race rating'
  )
```

## Legend-only helpers

Sometimes the table is already colored, or the color mapping is handled elsewhere. In that case, the `gtscale_color_*()` helpers attach only the legend.

```{r}
exibble |>
  gt() |>
  gt::data_color(
    columns = num,
    method = 'numeric',
    palette = c('#A0442C', 'white', '#0063B1')
  ) |>
  gtscale_color_continuous(
    column = num,
    palette = c('#A0442C', 'white', '#0063B1'),
    title = 'Numeric scale'
  )
```

## Scale specifications

For more control, use a `gtscale_spec`. Specs separate scale definition from application and legend placement.

```{r}
spec <- gtscale_spec_continuous(
  num,
  palette = c('#A0442C', 'white', '#0063B1'),
  labels = big_number_labels,
  width = '220px',
  title = 'Numeric scale'
) |>
  gtscale_spec_set_application(apply_to = 'fill') |>
  gtscale_spec_set_legend(placement = 'subtitle')

exibble |>
  gt() |>
  gtscale_apply_legend(spec)
```

This becomes more useful when the same scale needs to be reused or when you want to separate coloring from legend placement.

## Working with scales

`gtscales` is designed to accept the same kinds of helpers you would already use in plots.

### Labels and breaks

You can pass label functions and break functions directly.

```{r}
data.frame(share = c(0.12, 0.33, 0.57, 0.91)) |>
  gt() |>
  gtscale_data_color_bins(
    column = share,
    palette = c('#f7fbff', '#08306b'),
    bins = c(0, 0.25, 0.5, 0.75, 1),
    labels = label_percent(),
    title = 'Share bins'
  )
```

### Palette functions

Palette functions from `scales` can be supplied directly.

```{r}
data.frame(value = c(1, 10, 100, 1000)) |>
  gt() |>
  gtscale_data_color_continuous(
    column = value,
    palette = pal_viridis(),
    transform = 'log10',
    breaks = breaks_log(),
    labels = label_number(),
    title = 'Log scale'
  )
```

### Date and time data

Date-like columns work through the continuous and binned workflows. In many cases, `gtscales` can infer the appropriate transform from the column class.

```{r}
data.frame(
  when = as.Date(c('2024-01-01', '2024-01-20', '2024-02-10', '2024-03-05')),
  value = c(10, 18, 35, 52)
) |>
  gt() |>
  gtscale_data_color_bins(
    column = when,
    palette = pal_viridis(),
    bins = breaks_width('1 month'),
    labels = date_labels,
    width = '220px',
    title = 'Monthly bins'
  )
```

## Legend placement

Legends can be attached as source notes, subtitles, or titles.

Source notes are the default and are the most portable across outputs.

```{r}
exibble |>
  gt() |>
  gtscale_legend(
    gtscale_spec_continuous(
      num,
      palette = c('#A0442C', 'white', '#0063B1'),
      labels = big_number_labels,
      width = '220px',
      title = 'Numeric scale'
    ) |>
      gtscale_spec_set_legend(placement = 'source_note')
  )
```

When you want the legend closer to the heading, use `subtitle` or `title`.

```{r}
exibble |>
  gt() |>
  gtscale_legend(
    gtscale_spec_quantiles(
      num,
      palette = c('#fdd49e', '#fdbb84', '#ef6548', '#990000'),
      quantiles = 4,
      labels = big_number_labels,
      width = '220px',
      title = 'Quartiles'
    ) |>
      gtscale_spec_set_legend(placement = 'subtitle')
  )
```

```{r}
exibble |>
  gt() |>
  gtscale_legend(
    gtscale_spec_continuous(
      num,
      palette = c('#A0442C', 'white', '#0063B1'),
      labels = big_number_labels,
      width = '220px',
      title = 'Numeric scale'
    ) |>
      gtscale_spec_set_legend(placement = 'title')
  )
```

If you attach multiple legends to the same heading area, use `layout = "inline"` to place them side by side.

```{r}
data.frame(a = 1:3, b = 4:6) |>
  gt() |>
  gtscale_legend(
    gtscale_spec_continuous(
      a,
      palette = c('#f7fbff', '#08306b'),
      title = 'A'
    ) |>
      gtscale_spec_set_legend(placement = 'subtitle', layout = 'inline')
  ) |>
  gtscale_legend(
    gtscale_spec_continuous(
      b,
      palette = c('#fff5eb', '#7f2704'),
      title = 'B'
    ) |>
      gtscale_spec_set_legend(placement = 'subtitle', layout = 'inline')
  )
```

## Output support

The most mature output path is still `gt` HTML. The package also has validated workflows for LaTeX/PDF and Typst through the example files in `inst/examples`.

Current boundaries:

- HTML through `gt` is supported.
- LaTeX and Quarto PDF are supported through contextual legends.
- Typst is supported through the render-only workflow and example files.
- Word is not currently supported.

That means the safest general recommendation is to use source-note placement when you want the most consistent behavior across outputs, and then use heading placement when you control the rendering path more tightly.
